\documentclass[11pt,letterpaper]{article}

% ── Packages ──
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mathtools}
\usepackage{enumitem}
\usepackage{booktabs}
\usepackage{hyperref}
\usepackage{xcolor}
\usepackage{tcolorbox}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{tabularx}
\usepackage{array}

% ── Styling ──
\hypersetup{colorlinks=true, linkcolor=blue!60!black, citecolor=green!50!black, urlcolor=blue!70!black}
\pagestyle{fancy}
\fancyhf{}
\rhead{\small QuakeFloat8}
\lhead{\small February 2026}
\cfoot{\thepage}

\titleformat{\section}{\Large\bfseries\color{blue!60!black}}{}{0em}{}[\vspace{-0.5em}\rule{\textwidth}{0.4pt}]
\titleformat{\subsection}{\large\bfseries}{}{0em}{}
\titleformat{\subsubsection}{\normalsize\bfseries}{}{0em}{}

% ── Theorem Environments ──
\newtheorem{theorem}{Theorem}[section]
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{conjecture}[theorem]{Conjecture}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\theoremstyle{remark}
\newtheorem*{remark}{Remark}

% ── Boxes ──
\newtcolorbox{keyresult}{colback=green!5!white, colframe=green!50!black,
  title={\textbf{Key Result}}, fonttitle=\bfseries}


% ── Commands ──
\newcommand{\NMSE}{\mathrm{NMSE}}
\newcommand{\MSRE}{\mathrm{MSRE}}
\newcommand{\MSE}{\mathrm{MSE}}
\newcommand{\SQNR}{\mathrm{SQNR}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\R}{\mathbb{R}}
\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

% ── Document ──
\begin{document}

\begin{center}
  {\LARGE\bfseries QuakeFloat8 (QF8)} \\[6pt]
  {\Large A Provably Optimal Log-Domain 8-Bit Format} \\[4pt]
  {\Large for ML Multiply-Accumulate} \\[12pt]
  {\large February 2026} \\[4pt]
  {\normalsize Noah Everett}
\end{center}

\vspace{1em}

\begin{abstract}
QuakeFloat8 (QF8) is a block-scaled logarithmic 8-bit number format for machine learning 
in which multiplication reduces to 7-bit integer addition.
The format uses a 1-bit sign, a 7-bit \texttt{u3.4} fixed-point $\log_2$ code per element,
and a shared E8M0 block exponent per 32 elements.
This document consolidates all theoretical results, hardware cost analysis, training experiments,
and literature positioning for the QF8 project.

Key results:
(1)~Log-uniform quantization is provably minimax-optimal for NMSE over all source distributions (Theorem~\ref{thm:minimax}).
(2)~QF8 achieves $+6.6$~dB SQNR over IEEE FP8 E4M3 at the same effective storage cost.
(3)~The QF8 multiplier costs ${\sim}66$ gates vs.\ ${\sim}231$ for FP8 ($3.5\times$ cheaper).
(4)~TinyGPT-2 training with STE-based QAT matches FP32 validation loss.
\end{abstract}

\tableofcontents
\newpage

%% ═══════════════════════════════════════════════════════
\section{Introduction and Motivation}
\label{sec:intro}
%% ═══════════════════════════════════════════════════════

\subsection{The 8-Bit Frontier}

The push toward low-precision arithmetic for deep learning has been remarkably successful.
FP16 training is standard practice; BFloat16 is ubiquitous; and 8-bit formats (INT8, FP8 E4M3/E5M2)
are deployed in production hardware (NVIDIA H100, AMD MI300, Google TPU).
The industry consensus, codified in the OCP Microscaling (MX) specification, is that
8~bits per element with block scaling is sufficient for both training and inference
of large language models.

Yet a gap remains between the precision delivered by current 8-bit formats and what is theoretically
achievable. FP8 E4M3 provides 8~representable levels per octave (factor of~2).
Is this optimal? Can we do better at the same bit budget?

\subsection{The Quake Insight}

In 1999, the Quake III source code revealed a remarkable hack: the integer bit pattern of an IEEE~754
float is approximately a linear function of $\log_2(x)$. Reinterpreting float bits as integers,
and vice versa, allows cheap approximate logarithmic arithmetic.

QF8 makes this accident intentional. By storing values as fixed-point $\log_2$ codes, multiplication
becomes integer addition --- replacing an $O(n^2)$-gate multiplier with an $O(n)$-gate adder.
The format is paired with MX-style block scaling (E8M0 shared exponent per 32 elements)
to achieve full FP32 dynamic range.

\subsection{Contributions}

\begin{enumerate}[leftmargin=2em]
  \item \textbf{Formal optimality.} We prove the \emph{equalization property} of log-uniform
    quantization: $\MSRE = \varepsilon^2/12$ for \emph{any} source density, because the $x^2$
    factors in the NMSE ratio cancel exactly. This density-independence implies that log-uniform
    quantization is the unique minimax-optimal quantizer for NMSE over all source distributions
    (Theorem~\ref{thm:minimax}), complementing Sun \& Goyal's~[16] variable-rate optimality result
    and extending it to the fixed-rate minimax setting. Log-uniform quantization also minimizes
    worst-case product error for multiply-accumulate (Theorem~\ref{thm:product-optimal}).
  \item \textbf{Format design.} QF8 combines a \texttt{u3.4} fixed-point log code (16 levels per octave)
    with E8M0 block scaling, achieving $2\times$ the precision per octave of FP8 E4M3
    at the same effective storage cost (8.25 bits/element).
  \item \textbf{Hardware analysis.} Detailed gate-level models show the QF8 multiplier at ${\sim}66$~gates
    vs.\ ${\sim}231$ for FP8 mantissa multiplication ($3.5\times$ cheaper). The complete QF8-Medium MAC
    unit achieves near-INT8 area and power while providing floating-point dynamic range.
  \item \textbf{Training validation.} STE-based quantization-aware training on TinyGPT-2 shows QF8
    matching FP32 validation loss ($2.5445$ vs.\ $2.5450$), outperforming FP8 E4M3 ($2.5478$).
\end{enumerate}


%% ═══════════════════════════════════════════════════════
\section{The QF8 Format}
\label{sec:format}
%% ═══════════════════════════════════════════════════════

\subsection{Encoding Scheme}

\begin{definition}[QF8 Encoding]
A QF8-encoded block of $k = 32$ values consists of:
\begin{itemize}[leftmargin=2em]
  \item \textbf{Shared block scale} (8 bits): An E8M0 power-of-2 exponent $s \in \{2^{-127}, \ldots, 2^{127}\}$.
  \item \textbf{Per-element code} (8 bits each): 1~sign bit $\sigma \in \{0,1\}$ and a 7-bit
    unsigned \texttt{u3.4} fixed-point code $c \in \{0, 1, \ldots, 127\}$.
\end{itemize}
The reconstruction formula is:
\[
  \hat{x} = (-1)^\sigma \cdot s \cdot 2^{(c - 64)/16}, \qquad c \neq 0
\]
with $c = 0$ encoding exact zero. The effective storage cost is $8 + 8/32 = 8.25$~bits per element.
\end{definition}

\subsection{Logarithmic Spacing}

The \texttt{u3.4} code provides 3~integer bits and 4~fractional bits of $\log_2|x/s|$,
spanning 8~octaves with \textbf{16 representable levels per octave}.
This is the \emph{log-uniform quantizer}: the ratio between consecutive representable values is
constant at $r = 2^{1/16} \approx 1.0443$, giving a maximum relative quantization error of
\[
  \frac{r - 1}{2} \approx 2^{1/16} - 1 \approx 4.43\%.
\]
Compare FP8 E4M3, which has 8~levels per octave (3-bit mantissa) and a maximum relative error
of $1/16 = 6.25\%$ within each binade.

\subsection{Multiplication as Integer Addition}

For two QF8 values with log codes $c_a, c_b$:
\begin{align}
  \text{sign}_{\text{product}} &= \sigma_a \oplus \sigma_b \label{eq:sign-xor} \\
  \text{log-code}_{\text{product}} &= c_a + c_b \label{eq:log-add}
\end{align}
The product's log code is a simple 7-bit integer addition. In hardware, this replaces the
$4 \times 4$ mantissa multiplier needed for FP8 E4M3 with a 7-bit carry-lookahead adder ---
a ${\sim}3.5\times$ reduction in gate count (66 vs.\ 231 gates).

\subsection{Accumulation Strategy}

The classical weakness of log-domain arithmetic is that \emph{addition} in the original domain is expensive.
QF8 adopts the \textbf{Kulisch accumulation} strategy (Johnson, 2018):
\begin{enumerate}[leftmargin=2em]
  \item Multiply in log domain (integer addition of codes).
  \item Convert each product to linear domain via a tiny $2^4 = 16$-entry lookup table.
  \item Accumulate in a wide fixed-point register (32--48 bits for practical designs).
\end{enumerate}
This hybrid log-multiply / linear-accumulate approach gets cheap multiplies \emph{and} exact (or near-exact)
accumulation.

\subsection{BMD Realizability}

\begin{definition}[Bit-Manipulation Decodable (BMD)]
A decoder $\hat{D}: \{0,1\}^B \to \R$ is \emph{bit-manipulation decodable} if it can be expressed
as a fixed-length straight-line program over $\{+, -, \times, \gg, \ll, \mathbin{\&}, |, \oplus\}$
on machine-word integers, with the output reinterpreted as a float via type punning.
\end{definition}

\begin{theorem}[BMD Realizability]
\label{thm:bmd}
The log-uniform decode function $\hat{D}(n) = a \cdot r^n$ can be implemented via:
\begin{enumerate}[leftmargin=2em]
  \item \textbf{Lookup table:} $O(N)$ storage, $O(1)$ read.
  \item \textbf{Reinterpret cast (Quake trick):} $I = \alpha n + \beta$, then float reinterpret.
    Cost: 1~multiply + 1~add + type pun $= 3$~ops.
  \item \textbf{Quadratic approximation:} $2^{e+f} = 2^e \cdot (1 + 0.6602f + 0.3398f^2)$.
    Cost: 2~multiplies + 2~adds + 1~shift $= 5$~ops.
    Maximum relative error: $0.27\%$ (minimax optimal for constrained quadratic).
\end{enumerate}
\end{theorem}


%% ═══════════════════════════════════════════════════════
\section{Theoretical Results: Minimax Optimality}
\label{sec:minimax}
%% ═══════════════════════════════════════════════════════

\subsection{Notation and Setup}

Let $X$ be a positive random variable with pdf $f$ supported on $[a,b] \subset \R_{>0}$.
A $B$-bit scalar quantizer is a map $Q: [a,b] \to \hat{\mathcal{X}}$ with
$N = 2^B$ reconstruction points partitioning $[a,b]$ into cells $\{S_k\}_{k=0}^{N-1}$
with codepoints $\{c_k\}$.

\begin{definition}[Distortion Measures]
\label{def:distortion}
\begin{align}
  \delta_X &= X - Q(X), & \varepsilon_X &= \delta_X / X \\
  \MSE_X &= \E[\delta_X^2], & \NMSE_X &= \frac{\E[\delta_X^2]}{\E[X^2]} \label{eq:nmse-def} \\
  \MSRE_X &= \E[\varepsilon_X^2], & \SQNR &= 10\log_{10}(1/\NMSE) \text{ dB} \label{eq:sqnr-def}
\end{align}
\end{definition}

\begin{definition}[Auxiliary Quantities]
\label{def:auxiliary}
\begin{align}
  \alpha_X &= \E[X\delta_X]/\E[X^2] & &\text{(signal-error correlation)} \\
  \beta_X &= \E[Q(X)\delta_X]/\E[X^2] & &\text{(reconstruction-error correlation)} \\
  \gamma_X &= \E[Q(X)^2]/\E[X^2] & &\text{(quantized power ratio)} \\
  \rho_X &= \E[X \cdot Q(X)]/\E[X^2] & &\text{(signal-quantized correlation)}
\end{align}
Algebraic identities (no assumptions required):
\begin{equation}
  \alpha_X = \beta_X + \NMSE_X, \qquad
  \gamma_X = 1 - 2\alpha_X + \NMSE_X, \qquad
  \rho_X = 1 - \alpha_X
  \label{eq:algebraic-id}
\end{equation}
\end{definition}

\begin{definition}[Centroid Condition]
\label{def:centroid}
A quantizer satisfies the centroid condition if $\E[\delta_X \mid X \in S_k] = 0$ for every cell.
This holds for Lloyd-Max quantizers and approximately for any well-designed quantizer
in the high-resolution regime. Under the centroid condition:
$\beta_X = 0$, $\alpha_X = \NMSE_X$, and $\gamma_X = 1 - \NMSE_X$.
\end{definition}

\begin{definition}[Log-Uniform Quantizer]
\label{def:log-uniform}
The log-uniform quantizer with $N$ levels over dynamic range $[a,b]$
has cell boundaries $a \cdot r^k$ for $k = 0, \ldots, N$, where
$r = (b/a)^{1/N} = 2^{R/N}$ and $R = \log_2(b/a)$ is the dynamic range in octaves.
With geometric midpoint codepoints $c_k = a \cdot r^{k+1/2}$,
the relative cell width is constant:
\begin{equation}
  \frac{w(x)}{x} = r - 1 = e^\varepsilon - 1 \approx \varepsilon,
  \qquad \varepsilon = \frac{R \ln 2}{N}
  \label{eq:rel-cell-width}
\end{equation}
\end{definition}

\subsection{Relation to Classical Quantization Theory}
\label{subsec:prior-theory}

The classical companding literature studies optimal nonuniform quantizers under MSE distortion.
Bennett~[13] established the foundational high-resolution framework: for a compander with local step size
$w(x)$ and source pdf $f$, the distortion is $D = \frac{1}{12}\int w(x)^2 f(x)\,dx$, giving the
$\varepsilon^2/12$ approximation for uniform quantizers.
Panter \& Dite~[14] derived the MSE-optimal compander (the $p(x)^{1/3}$ one-third-power rule): it achieves
equal MSE contribution per cell, but depends critically on the source distribution.
The standard textbook treatments~[17, 18] cover these results thoroughly.

For \emph{relative} error metrics, Smith~[15] observed that logarithmic companding achieves approximately
amplitude-independent SNR for PCM telephony --- meaning the SNR is roughly constant as a function of
\emph{signal level} (e.g., sinusoidal amplitude). This is structurally distinct from the equalization
property of Theorem~\ref{thm:minimax}: Smith's result is about signal amplitude, not about source
probability density. Indeed, Smith's distortion formula retains a distribution-dependent factor
(e.g., $0.707$ for exponential, $0.798$ for Gaussian distributions), so the distortion \emph{is}
distribution-dependent in his framework.

Most directly related, Sun \& Goyal~[16] study scalar quantization explicitly under the MSRE metric.
They prove that logarithmic companding is \emph{variable-rate}-optimal for MSRE: given a source
distribution $f$, the entropy-constrained optimal quantizer is logarithmic. However, they also
show that logarithmic companding is \emph{not} generally optimal for fixed-rate quantization ---
the optimal fixed-rate MSRE quantizer depends on the source distribution.

Theorem~\ref{thm:minimax} provides a complementary and stronger fixed-rate result:
the log-uniform quantizer achieves MSRE $= \varepsilon^2/12$ for \emph{any} source density
(the equalization property), and this constant value is the fixed-rate minimax optimum.
The key structural insight is the $x^2$ cancellation in the NMSE ratio, which is absent in all
prior companding results: it makes the log-uniform quantizer's error density-independent rather than
merely amplitude-independent. The minimax framing is not addressed in any prior work found.

\textbf{Why minimax for ML?}
The ML setting motivates worst-case (minimax) rather than average-case (Bayes) analysis for two reasons.
First, weight and activation distributions shift substantially across layers, training stages, and models,
so a quantizer optimal for one distribution may be poor for another.
Second, the equalization property has direct practical value: a format with density-independent NMSE
requires no per-layer calibration of the quantizer grid, and its error bound holds regardless of whether
the layer distribution matches any assumed prior.
The log-uniform quantizer's minimax value $\varepsilon^2/12$ is not just a worst-case bound ---
by the equalization property, it is achieved for every source distribution, including Gaussian, uniform,
Laplace, and the empirical distributions observed in trained networks.

\subsection{Theorem 2.1: Minimax NMSE Optimality}

\begin{theorem}[Minimax NMSE Optimality of Log-Uniform Quantization]
\label{thm:minimax}
Among all $N$-level quantizers on $[a,b]$, the log-uniform quantizer uniquely minimizes the worst-case
NMSE (and MSRE) over all densities:
\[
  Q^*_{\log} = \argmin_{Q \in \mathcal{Q}_N}\; \max_{f \in \mathcal{F}[a,b]}\; \NMSE(Q, f)
\]
The minimax value is:
\begin{equation}
  \boxed{\NMSE^* = \frac{\varepsilon^2}{12} + O(\varepsilon^4) = \frac{(R \ln 2)^2}{12 N^2}}
  \label{eq:minimax-nmse}
\end{equation}
For $B = 8$ bits, $R = 16$ octaves: $\NMSE^* = 1.564 \times 10^{-4}$ ($\SQNR = 38.1$~dB).
\end{theorem}

\begin{proof}
\textbf{Step 1 (Density-independence).}
For the log-uniform quantizer, $w(x) = x \cdot \varepsilon + O(x\varepsilon^2)$.
Under the high-resolution approximation:
\[
  \NMSE = \frac{\int_a^b \frac{w(x)^2}{12} f(x)\,dx}{\int_a^b x^2 f(x)\,dx}
  = \frac{\frac{\varepsilon^2}{12}\int_a^b x^2 f(x)\,dx}{\int_a^b x^2 f(x)\,dx}
  = \frac{\varepsilon^2}{12}
\]
The $x^2$ factors cancel exactly. Similarly, $\MSRE = \varepsilon^2/12$ for any density $f$.
This is the \emph{equalization property} of log-uniform quantization.

This is confirmed by exhaustive computation over three test densities (log-uniform, uniform, $f \propto x^2$)
with $N = 256$, $R = 16$: all give $\NMSE = 1.564 \times 10^{-4}$ within $< 0.1\%$ deviation.

\textbf{Step 2 (Non-constant $\phi$ is worse for some density).}
Define $\phi(x) = w(x)/x$ (relative cell width). For any non-log-uniform quantizer,
$\phi$ is not constant. Let $x^* = \argmax_x \phi(x)$ and $\phi^* = \phi(x^*)$.

We first show $\phi^* > \varepsilon$. Any $N$-cell quantizer on $[a,b]$ satisfies the
coverage constraint $\int_a^b dx/w(x) = N$, i.e.\
\begin{equation}
  \int_a^b \frac{dx}{\phi(x)\,x} = N.
  \label{eq:coverage}
\end{equation}
For the log-uniform quantizer $\phi \equiv \varepsilon$, this gives $(1/\varepsilon)\ln(b/a) = N$ since
$\varepsilon = R\ln 2/N$ and $\ln(b/a) = R\ln 2$. Suppose for contradiction that $\phi$
is non-constant but $\phi(x) \le \varepsilon$ everywhere. Then $1/\phi(x) \ge 1/\varepsilon$
everywhere, with strict inequality on a set of positive measure, so
\[
  N = \int_a^b \frac{dx}{\phi(x)\,x} > \int_a^b \frac{dx}{\varepsilon\,x}
  = \frac{\ln(b/a)}{\varepsilon} = N,
\]
a contradiction. Therefore $\phi^* > \varepsilon$ for any non-log-uniform quantizer.

For any $\eta > 0$, the uniform density on $[x^* - \eta, x^* + \eta]$ achieves:
\[
  \MSRE \geq \frac{\phi(x^*)^2}{12} - g(\eta), \qquad g(\eta) \to 0 \text{ as } \eta \to 0
\]
Therefore $\max_f \MSRE(Q, f) \geq (\phi^*)^2/12 > \varepsilon^2/12 = \MSRE^*$.

\textbf{Step 3 (Uniqueness).}
If $Q \neq Q_{\log}$, then $\phi$ is non-constant and Step~2 gives strict inequality. \qed
\end{proof}

\begin{corollary}[Minimax MSRE]
\label{cor:minimax-msre}
The same quantizer also minimizes worst-case MSRE. The proof is identical, with
$w(x)^2 f(x)/x^2$ replacing $w(x)^2 f(x)/\E[X^2]$.
\end{corollary}


%% ═══════════════════════════════════════════════════════
\section{Product Error Decomposition}
\label{sec:product}
%% ═══════════════════════════════════════════════════════

This section addresses how quantization error propagates through multiplication ---
the core operation in neural network inference.

\subsection{The General Formula}

\begin{theorem}[General Product Error Decomposition]
\label{thm:product-general}
Let $X, Y$ be independent positive random variables with quantizers $Q_X, Q_Y$.
Then:
\begin{equation}
  \boxed{\NMSE_{\mathrm{prod}} = \NMSE_X + \NMSE_Y + \NMSE_X \cdot \NMSE_Y + 2\alpha_X \alpha_Y
  - 2\alpha_X \cdot \NMSE_Y - 2\NMSE_X \cdot \alpha_Y}
  \label{eq:product-general}
\end{equation}
where $\alpha_X = \E[X\delta_X]/\E[X^2]$.
Equivalently:
\begin{equation}
  \NMSE_{\mathrm{prod}} = 1 - 2\rho_X \rho_Y + \gamma_X \gamma_Y
  \label{eq:product-rho-gamma}
\end{equation}
This is exact, using only $X \perp Y$. No high-resolution, centroid, or distributional assumptions.
\end{theorem}

\begin{proof}
\textbf{Step 1.} Decompose the product error:
\[
  XY - Q_X Q_Y = (Q_X + \delta_X)(Q_Y + \delta_Y) - Q_X Q_Y
  = \underbrace{Q_X \delta_Y}_{A} + \underbrace{\delta_X Q_Y}_{B} + \underbrace{\delta_X \delta_Y}_{C}
\]

\textbf{Step 2.} Expand $\E[(A+B+C)^2]$ using $X \perp Y$. Since $Q_X, \delta_X$ depend only
on~$X$ and $Q_Y, \delta_Y$ depend only on~$Y$, every product of an $X$-function and a $Y$-function
factors:
\begin{align*}
  \E[(A+B+C)^2] &= \E[Q_X^2]\E[\delta_Y^2] + \E[Q_Y^2]\E[\delta_X^2] + \E[\delta_X^2]\E[\delta_Y^2] \\
  &\quad + 2\E[Q_X\delta_X]\E[Q_Y\delta_Y] + 2\E[Q_X\delta_X]\E[\delta_Y^2]
  + 2\E[\delta_X^2]\E[Q_Y\delta_Y]
\end{align*}

\textbf{Step 3.} Divide by $\E[X^2]\E[Y^2]$ and substitute
$\beta_X = \alpha_X - \NMSE_X$, $\gamma_X = 1 - 2\alpha_X + \NMSE_X$:
\[
  \NMSE_{\mathrm{prod}} = \gamma_X \NMSE_Y + \gamma_Y \NMSE_X + \NMSE_X \NMSE_Y
  + 2\beta_X \beta_Y + 2\beta_X \NMSE_Y + 2\NMSE_X \beta_Y
\]
Expanding $\beta = \alpha - \NMSE$ and simplifying yields~\eqref{eq:product-general}. \qed
\end{proof}

\subsection{Centroid Simplification}

\begin{corollary}[Product Error Under Centroid Condition]
\label{cor:product-centroid}
If both quantizers satisfy the centroid condition ($\alpha_X = \NMSE_X$, $\alpha_Y = \NMSE_Y$), then:
\begin{equation}
  \boxed{1 - \NMSE_{\mathrm{prod}} = (1 - \NMSE_X)(1 - \NMSE_Y)}
  \label{eq:product-preservation}
\end{equation}
Equivalently: $\NMSE_{\mathrm{prod}} = \NMSE_X + \NMSE_Y - \NMSE_X \cdot \NMSE_Y$.
\end{corollary}

\begin{proof}
Substitute $\alpha_X = n_X$, $\alpha_Y = n_Y$ into~\eqref{eq:product-general}:
$n_X + n_Y + n_X n_Y + 2n_X n_Y - 2n_X n_Y - 2n_X n_Y = n_X + n_Y - n_X n_Y$. \qed
\end{proof}

\begin{remark}[Sign of the Cross-Term]
The correct sign is \textbf{minus} ($n_X + n_Y - n_X n_Y$), not plus as in the original formulation.
The identity $(1 - n_X)(1 - n_Y) = 1 - n_{\mathrm{prod}}$ has a clean interpretation:
the signal preservation fraction of the product equals the product of individual preservation fractions.
However, since $n_X n_Y = O(1/N^4)$ while $n_X + n_Y = O(1/N^2)$,
the distinction is negligible in practice.
The corrected sign and the equivalence of the two forms~\eqref{eq:product-general} and~\eqref{eq:product-rho-gamma}
were confirmed by exhaustive computation over 14,641 integer grid points covering all combinations
of $(\NMSE_X, \NMSE_Y) \in \{0, 0.01, \ldots, 1\}^2$.
\end{remark}

\begin{remark}[Correction from Original]
The original Theorem 3.1 used an MSRE-based formula with an invalid factorization
($\E[X^2 \varepsilon_X^2] = \E[X^2]\E[\varepsilon_X^2]$, which fails because
$\varepsilon_X$ depends deterministically on~$X$).
For non-log-uniform quantizers, the MSRE formula can be off by factors of $6$--$1000\times$.
The corrected formula~\eqref{eq:product-general} uses only the valid factorization $X \perp Y$.
\end{remark}

\subsection{Specialization to Log-Uniform Quantizers}

\begin{theorem}[Log-Uniform Product Error]
\label{thm:product-loguniform}
For log-uniform quantizers with geometric midpoints, $\MSRE = \NMSE = \varepsilon^2/12$
for any density, and:
\begin{equation}
  \NMSE_{\mathrm{prod}} = \frac{(R\ln 2)^2}{6N^2} + O(1/N^4) = 2\NMSE^* + O(1/N^4)
\end{equation}
The original MSRE-based formula gives the correct answer to leading order for log-uniform quantizers
(despite being invalid in general) because the equalization property makes
$\MSRE \approx \NMSE$ with ratio $1.0009$ for $N = 16$.
\end{theorem}

\subsection{Optimality for Multiplication}

\begin{theorem}[Log-Uniform Optimal for Multiplication]
\label{thm:product-optimal}
Among all $N$-level quantizer pairs $(Q_X, Q_Y)$ on $[a,b]$, the log-uniform quantizer minimizes
the worst-case product NMSE:
\[
  Q_X^* = Q_Y^* = Q_{\log} = \argmin_{Q_X, Q_Y \in \mathcal{Q}_N}
  \max_{f_X, f_Y} \NMSE_{\mathrm{prod}}
\]
The minimax value is $2n^* - (n^*)^2 = 1 - (1 - n^*)^2$ where $n^* = \varepsilon^2/12$.
\end{theorem}

\begin{proof}
Uses two properties:
(1)~\emph{Monotonicity}: under the centroid condition, $\partial/\partial n_X(n_X + n_Y - n_X n_Y) = 1 - n_Y > 0$.
(2)~\emph{Density-independence}: $Q_{\log}$ achieves $\NMSE = n^*$ for all densities.
For any $Q_X \neq Q_{\log}$, Theorem~\ref{thm:minimax} provides $f_X^*$ with $\NMSE(Q_X, f_X^*) > n^*$,
giving strictly worse product NMSE.
\end{proof}

\subsection{Extension to Dot Products}

\begin{corollary}[Dimension-Free Dot Product NMSE]
\label{cor:dotproduct}
For $Z = \sum_{i=1}^d w_i x_i$ with independent pairs, iid within type,
and centroid-condition quantizers:
\begin{equation}
  \NMSE_Z \approx \NMSE_w + \NMSE_x + O(1/N^4)
\end{equation}
The NMSE \emph{does not grow with dimension~$d$}.
\end{corollary}

\begin{proof}[Proof sketch]
Under independence across dimensions:
$\E[(Z - \hat{Z})^2] = d \cdot \E[w^2 x^2](n_w + n_x - n_w n_x)$
and $\E[Z^2] = d \cdot \E[w^2 x^2]$. Division yields the result.
\end{proof}


%% ═══════════════════════════════════════════════════════
\section{Exact MSRE of Log-Uniform Quantization}
\label{sec:exact-msre}
%% ═══════════════════════════════════════════════════════

\begin{theorem}[Exact Per-Cell MSRE]
\label{thm:exact-msre}
For a log-uniform quantizer cell with common ratio $r = 2^{R/N}$ and geometric midpoint codepoint,
the exact MSRE (under uniform intra-cell density) is:
\begin{equation}
  \MSRE_k = 2 - \frac{2\sqrt{r}\,\ln r}{r - 1}
  \label{eq:exact-msre}
\end{equation}
For small $\varepsilon = \ln r = R\ln 2/N$:
\begin{equation}
  \boxed{\MSRE_k = \frac{\varepsilon^2}{12} - \frac{7\varepsilon^4}{2880} + O(\varepsilon^6)}
  \label{eq:msre-taylor}
\end{equation}
\end{theorem}

\begin{proof}
For a cell $[a, ra)$ with geometric midpoint $c = a\sqrt{r}$, substituting $u = x/a$:
\[
  \MSRE_k = \frac{1}{r-1}\int_1^r \!\left(1 - \frac{\sqrt{r}}{u}\right)^{\!2} du
  = \frac{2(r-1) - 2\sqrt{r}\ln r}{r - 1}
  = 2 - \frac{2\sqrt{r}\,\ln r}{r - 1}
\]
For the Taylor expansion, substitute $r = e^\varepsilon$:
\[
  \frac{2\sqrt{r}\,\ln r}{r-1} = \frac{2\varepsilon\, e^{\varepsilon/2}}{e^\varepsilon - 1}
  = \frac{2(1 + \varepsilon/2 + \varepsilon^2/8 + \cdots)}{1 + \varepsilon/2 + \varepsilon^2/6 + \cdots}
  = 2\!\left(1 - \frac{\varepsilon^2}{24} + O(\varepsilon^3)\right)
\]
Therefore $\MSRE_k = 2 - (2 - \varepsilon^2/12) = \varepsilon^2/12 + O(\varepsilon^4)$.
This \emph{confirms} the high-resolution approximation. \qed
\end{proof}

\begin{remark}[Correction of $\varepsilon^2/2$ Error]
The original derivation approximated $e^\varepsilon - 1 \approx \varepsilon$,
omitting the $\varepsilon^2/2$ correction in the denominator.
The correct denominator is $\varepsilon(1 + \varepsilon/2 + O(\varepsilon^2))$;
this extra factor converts the erroneous $\varepsilon^2/2$ to the correct $\varepsilon^2/12$ ---
a factor of 6 correction.
\end{remark}

\begin{table}[h]
\centering
\caption{Numerical verification: exact MSRE vs.\ approximations.}
\label{tab:msre-verify}
\begin{tabular}{@{}ccccc@{}}
\toprule
$\varepsilon$ & Exact MSRE & $\varepsilon^2/12$ & $\varepsilon^2/2$ (wrong) & Exact / $(\varepsilon^2/12)$ \\
\midrule
0.01  & $8.333 \times 10^{-6}$  & $8.333 \times 10^{-6}$  & $5.000 \times 10^{-5}$  & 1.000 \\
0.05  & $2.083 \times 10^{-4}$  & $2.083 \times 10^{-4}$  & $1.250 \times 10^{-3}$  & 0.9999 \\
0.10  & $8.331 \times 10^{-4}$  & $8.333 \times 10^{-4}$  & $5.000 \times 10^{-3}$  & 0.9997 \\
0.20  & $3.329 \times 10^{-3}$  & $3.333 \times 10^{-3}$  & $2.000 \times 10^{-2}$  & 0.9989 \\
\bottomrule
\end{tabular}
\end{table}


%% ═══════════════════════════════════════════════════════
\section{Separation Result: QF8 vs.\ FP8}
\label{sec:separation}
%% ═══════════════════════════════════════════════════════

\subsection{Exponential MSRE Separation}

\begin{theorem}[Exponential Separation Under MSRE]
\label{thm:separation}
For $X \sim \mathrm{LogNormal}(0, \sigma^2)$ quantized to $N$ levels:
\begin{enumerate}[label=(\roman*)]
  \item \textbf{Under MSRE:} the ratio of uniform quantization MSRE to logarithmic quantization MSRE
    is exponential in $\sigma^2$:
    \begin{equation}
      \frac{\MSRE_{\mathrm{uniform}}}{\MSRE_{\mathrm{log}}} = e^{\Theta(\sigma^2)}
      \label{eq:msre-separation}
    \end{equation}
  \item \textbf{Under MSE:} the two quantizers achieve comparable distortion (ratio $\approx 1$).
\end{enumerate}
\end{theorem}

This result is significant because MSRE is the natural metric for floating-point systems ---
it measures relative accuracy, which is what determines model quality.

\begin{table}[h]
\centering
\caption{MSRE and MSE ratios (uniform / log) for LogNormal($0, \sigma^2$), $N = 256$.}
\label{tab:separation}
\begin{tabular}{@{}ccccc@{}}
\toprule
$\sigma$ & MSE ratio & MSE dB & MSRE ratio & MSRE dB \\
\midrule
1.00  & 0.98  & $-0.1$ & 8.5        & 9.3  \\
1.50  & 0.99  & $-0.0$ & 1,520      & 31.8 \\
2.00  & 0.99  & $-0.0$ & 346,000    & 55.4 \\
2.77  & 1.00  & $0.0$  & $3.0 \times 10^9$ & 94.8 \\
\bottomrule
\end{tabular}
\end{table}

\begin{proof}[Proof sketch]
For $X = e^Z$ with $Z \sim \mathcal{N}(0, \sigma^2)$:
Log quantization on $X$ corresponds to uniform quantization on $Z$ (Gaussian),
giving $\MSRE_{\log} \approx \varepsilon^2/12$.
Uniform quantization on $X$ translates to highly nonuniform quantization on $Z = \ln X$:
at $Z = k\sigma$, the local step in log-space grows as $\propto e^{2k\sigma}$,
and integrating against the Gaussian density yields $\MSRE_{\mathrm{uniform}} \propto e^{c\sigma^2}$.
MSE is insensitive because it is dominated by absolute error on large values
where both quantizers are comparable.
\end{proof}

\subsection{QF8 vs.\ FP8 E4M3: Detailed Comparison}

\begin{table}[h]
\centering
\caption{Within-binade precision comparison.}
\label{tab:qf8-vs-fp8-binade}
\begin{tabular}{@{}lcccc@{}}
\toprule
Metric & IEEE E4M3 & Log-uniform (8/oct) & Ratio & dB gap \\
\midrule
Average MSRE (per binade)  & $6.510 \times 10^{-4}$ & $6.256 \times 10^{-4}$ & 1.041 & 0.2 \\
Peak relative error        & 0.0625               & 0.04427              & 1.41  & ---   \\
Worst-case MSRE (HR)       & $1.302 \times 10^{-3}$ & $6.256 \times 10^{-4}$ & 2.08  & 3.18 \\
\bottomrule
\end{tabular}
\end{table}

\begin{keyresult}
QF8 achieves $\SQNR = 38.1$~dB vs.\ FP8 E4M3's 31.5~dB on $\mathcal{N}(0,1)$
--- a \textbf{6.6~dB advantage}, corresponding to a ${\sim}4.6\times$ reduction in quantization noise power.
This advantage is structural (16 vs.\ 8 levels per octave) and holds across all tested distributions.

\medskip
\begin{tabular}{@{}lcc@{}}
\toprule
Format & SQNR on $\mathcal{N}(0,1)$ & Gap from QF8 \\
\midrule
QF8 (log-uniform, 256 levels) & 38.1 dB & --- \\
FP8 E4M3 (IEEE)               & 31.5 dB & 6.6 dB worse \\
\bottomrule
\end{tabular}
\end{keyresult}

\begin{remark}
The FP8 E4M3 SQNR of 31.5~dB is verified by per-binade analysis across all 126 positive representable
values, correcting an earlier figure of 25.1~dB that arose from an error in the dynamic-range assumption.
\end{remark}


%% ═══════════════════════════════════════════════════════
\section{Hardware Cost Analysis}
\label{sec:hardware}
%% ═══════════════════════════════════════════════════════

This section presents gate-level hardware models anchored to published silicon data from
Horowitz (ISSCC 2014), Johnson (2018 ELMA ASIC at 28nm), and NVIDIA architecture whitepapers.

\subsection{Reference Data}

\textbf{Horowitz ISSCC 2014} (45nm CMOS): 8-bit integer multiply = 0.2 pJ,
8-bit add = 0.03 pJ, 32-bit FP multiply = 3.7 pJ.

\textbf{Johnson ELMA synthesis} (28nm):
8-bit ELMA/38-bit Kulisch = 0.96$\times$ power, 1.12$\times$ area vs.\ INT8/32-bit MAC.
16-bit ELMA = 0.59$\times$ power, 0.68$\times$ area vs.\ FP16 FMA.

\subsection{QF8 Multiply Path}

The core QF8 advantage: multiplication reduces to a 7-bit integer addition plus sign XOR.

\begin{table}[h]
\centering
\caption{QF8 multiply-path gate count.}
\label{tab:qf8-multiply}
\begin{tabular}{@{}lcc@{}}
\toprule
Component & Gates & Area ($\mu\text{m}^2$, 7nm) \\
\midrule
Sign XOR                    & 1   & 0.04 \\
7-bit CLA adder             & 50  & 2.0  \\
Overflow/saturation detect  & 5   & 0.2  \\
Zero detection              & 10  & 0.4  \\
\midrule
\textbf{Total: multiply}    & \textbf{66} & \textbf{2.6} \\
\bottomrule
\end{tabular}
\end{table}

Compare: INT8 $8\times 8$ multiplier $= 400$ gates; FP8 $4\times 4$ mantissa multiplier $+ $ exponent logic $= 230$ gates; FP16 $11\times 11$ multiplier $= 850$ gates; FP32 $24\times 24$ multiplier $= 3200$ gates.
The QF8 multiply path is \textbf{$3.5\times$ to $48\times$ cheaper} than alternatives.

\subsection{QF8-Medium MAC Unit}

The recommended design point: 6-bit barrel shifter and 32-bit carry-save Kulisch accumulator
for intra-block accumulation, with block-pair results reduced via FP32 arithmetic
(matching MX-format accelerator practice).

\begin{table}[h]
\centering
\caption{QF8-Medium MAC unit breakdown (7nm, 1 GHz).}
\label{tab:qf8-medium}
\begin{tabular}{@{}lcc@{}}
\toprule
Component & Gates & Area ($\mu\text{m}^2$) \\
\midrule
Multiply path (7-bit log adder)          & 50   & 2.0  \\
exp2 LUT ($16 \times 12$-bit ROM)        & 100  & 4.0  \\
6-bit barrel shifter                     & 150  & 6.0  \\
32-bit carry-save Kulisch accumulator    & 300  & 12.0 \\
Control/zero-detect/mux                  & 50   & 2.0  \\
\midrule
\textbf{Total}                           & \textbf{650} & \textbf{26.0} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Master Comparison}

\begin{table}[h]
\centering
\caption{Per-MAC unit comparison at 7nm, 1 GHz.}
\label{tab:master-comparison}
\begin{tabular}{@{}lccccc@{}}
\toprule
Metric & INT8 & FP8 E4M3 & FP16 & FP32 & \textbf{QF8-Med} \\
\midrule
Gate count              & 750   & 1,350 & 2,400  & 5,950  & \textbf{650}  \\
Area ($\mu\text{m}^2$)  & 30    & 54    & 96     & 238    & \textbf{26}   \\
Energy (pJ/MAC)         & 0.040 & 0.070 & 0.130  & 0.350  & \textbf{0.033}\\
Pipeline stages         & 2     & 3     & 3      & 3--4   & \textbf{2}    \\
Critical path (gates)   & 24--30& 35--45& 38--48 & 55--70 & \textbf{20--26}\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{QF8-Medium savings ratios vs.\ each baseline.}
\label{tab:savings-ratios}
\begin{tabular}{@{}lcccc@{}}
\toprule
Metric & vs FP32 & vs FP16 & vs FP8 & vs INT8 \\
\midrule
Area savings    & \textbf{89\%} & \textbf{73\%} & \textbf{52\%} & \textbf{13\%}  \\
Power savings   & \textbf{91\%} & \textbf{75\%} & \textbf{53\%} & \textbf{17\%}  \\
Latency improvement & 55--65\% & 35--45\% & 25--35\% & $\sim$5--10\% \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Systolic Array Scaling ($128 \times 128$)}

\begin{table}[h]
\centering
\caption{128$\times$128 systolic array estimates at 7nm, 1~GHz.}
\label{tab:systolic}
\begin{tabular}{@{}lccccc@{}}
\toprule
Metric & INT8 & FP8 & FP16 & FP32 & \textbf{QF8-Med} \\
\midrule
Compute area (mm$^2$) & 0.49 & 0.88 & 1.57 & 3.90 & \textbf{0.43} \\
Total power (W)        & 0.68 & 1.19 & 2.21 & 5.93 & \textbf{0.56} \\
TOPS/W                 & 48.2 & 27.6 & 14.8 & 5.5  & \textbf{58.5} \\
TOPS/mm$^2$            & 66.9 & 37.3 & 20.9 & 8.4  & \textbf{76.2} \\
\bottomrule
\end{tabular}
\end{table}

\subsection{The Break-Even Analysis}

The log-domain advantage scales differently with bit-width because the multiplier is $O(n^2)$
while the adder is $O(n)$:

\begin{table}[h]
\centering
\caption{Log-domain break-even across bit widths.}
\label{tab:breakeven}
\begin{tabular}{@{}ccccl@{}}
\toprule
Bit-width & Mult saved (gates) & LUT overhead (gates) & Net & vs INT-$N$ \\
\midrule
4   & 38     & 120  & $-82$     & $1.5\times$ worse \\
6   & 130    & 180  & $-50$     & $1.2\times$ worse \\
\textbf{8}   & \textbf{364}    & \textbf{300}  & $+64$     & \textbf{$0.92\times$ (slight win)} \\
12  & 1,145  & 500  & $+645$    & $0.63\times$ \\
16  & 2,925  & 800  & $+2{,}125$& $0.40\times$ \\
\bottomrule
\end{tabular}
\end{table}

\textbf{The crossover is at ${\sim}7$--$8$~bits.}
Below 7~bits, the LUT and accumulator overhead exceeds multiplier savings.
Above 8~bits, log-domain wins decisively and the advantage grows quadratically.
This confirms Johnson's 2018 finding that 16-bit ELMA achieves 41\% power reduction over FP16.

\subsection{The Honest Bottom Line}

\begin{itemize}[leftmargin=2em]
  \item \textbf{QF8 vs FP32/FP16:} Clear, large, robust win ($3$--$10\times$ savings).
  \item \textbf{QF8 vs FP8:} Clear win (${\sim}2\times$ area and power savings).
  \item \textbf{QF8 vs INT8:} Slight win ($0.87\times$ area, $0.83\times$ power).
    The carry-save Kulisch design tips the balance in QF8's favour.
  \item \textbf{The real QF8 value is $2\times$ precision-per-octave at INT8-equivalent
    hardware cost and half the cost of FP8.}
\end{itemize}


%% ═══════════════════════════════════════════════════════
\section{Training Results}
\label{sec:training}
%% ═══════════════════════════════════════════════════════

\subsection{TinyGPT-2 Experiment Configuration}

\begin{table}[h]
\centering
\caption{Training configuration.}
\label{tab:training-config}
\begin{tabular}{@{}ll@{}}
\toprule
Parameter & Value \\
\midrule
Model       & TinyGPT-2 ($d_{\text{model}} = 128$, heads $= 4$, layers $= 2$, $d_{\text{ff}} = 512$) \\
Vocabulary  & 256 (byte-level), $\text{seq\_len} = 128$ \\
Training    & 500 steps, batch $= 4$, lr $= 3 \times 10^{-4}$, cosine decay \\
Quantization& Block-scaled round-trip + STE, block\_size $= 32$ \\
Device      & CPU \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Results}

\begin{keyresult}
\begin{tabular}{@{}lcccc@{}}
\toprule
Model & Final Train & Final Val & $\Delta_{\text{val}}$ vs FP32 \\
\midrule
FP32       & 2.5388 & 2.5450 & --- \\
FP8 E4M3   & 2.5401 & 2.5478 & $+0.0028$ ($+0.11\%$) \\
\textbf{QF8} & \textbf{2.5388} & \textbf{2.5445} & $\mathbf{-0.0005}$ ($-0.02\%$) \\
\bottomrule
\end{tabular}

\medskip
QF8 matches FP32 validation loss to within noise ($-0.02\%$),
while FP8 E4M3 incurs a small penalty ($+0.11\%$).
\end{keyresult}

\subsection{Training Loss Trajectory}

Both QF8 and FP8 track the FP32 loss curve closely throughout training.
Selected validation loss checkpoints:

\begin{table}[h]
\centering
\caption{Validation loss at selected steps.}
\label{tab:val-loss}
\begin{tabular}{@{}cccc@{}}
\toprule
Step & FP32 & FP8 E4M3 & QF8 \\
\midrule
1    & 5.3532 & 5.3568 & 5.3520 \\
100  & 2.7647 & 2.7656 & 2.7645 \\
250  & 2.5749 & 2.5753 & 2.5753 \\
375  & 2.4425 & 2.4443 & 2.4430 \\
500  & 2.5450 & 2.5478 & 2.5445 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{Benchmark Results}

Random matrix multiplication accuracy (measured vs.\ float64 ground truth):

\begin{table}[h]
\centering
\caption{Matrix multiplication SQNR across sizes.}
\label{tab:matmul-sqnr}
\begin{tabular}{@{}lcccc@{}}
\toprule
Size & bfloat16 & float32 & FP8 E4M3 (block) & \textbf{QF8} \\
\midrule
$16 \times 32 \times 16$     & 52.3 dB & 139.0 dB & 28.6 dB & \textbf{35.3 dB} \\
$64 \times 128 \times 64$    & 52.5 dB & 133.6 dB & 28.4 dB & \textbf{35.1 dB} \\
$128 \times 256 \times 128$  & 52.6 dB & 130.8 dB & 28.5 dB & \textbf{35.1 dB} \\
\bottomrule
\end{tabular}
\end{table}

\begin{table}[h]
\centering
\caption{SQNR across value distributions.}
\label{tab:dist-sqnr}
\begin{tabular}{@{}lccc@{}}
\toprule
Distribution & FP8 E4M3 (block) & QF8 & QF8 advantage \\
\midrule
$\mathcal{N}(0, 0.02^2)$ --- weights    & 31.4 dB & 38.2 dB & $+$6.7 dB \\
$\mathcal{N}(0, 1)$ --- post-LN acts    & 31.6 dB & 37.9 dB & $+$6.3 dB \\
$\mathrm{LogN}(0, 1)$ --- gradients     & 31.5 dB & 38.3 dB & $+$6.8 dB \\
$\mathrm{Laplace}(0, 0.02)$ --- sparse  & 31.5 dB & 38.0 dB & $+$6.5 dB \\
90\% sparse $+$ $\mathcal{N}(0,1)$      & 31.7 dB & 38.3 dB & $+$6.6 dB \\
\bottomrule
\end{tabular}
\end{table}

The $+6.6$~dB advantage is consistent across all tested distributions and matrix sizes,
confirming that it is a structural property of the format rather than a distribution-specific artifact.


%% (Section 6 "Lean 4 Verification" removed — numerical checks folded into relevant theorem remarks)


%% ═══════════════════════════════════════════════════════
\section{Literature Positioning}
\label{sec:literature}
%% ═══════════════════════════════════════════════════════

\subsection{Relationship to Johnson (2018)}

The closest prior work is Johnson's ``Rethinking Floating Point for Deep Learning''
(arXiv:1811.01721), which proposed ELMA: an 8-bit format using log-domain multiply
with Kulisch accumulation. QF8's core mechanism --- log-domain multiply via integer addition
with linear accumulation --- \emph{is} Johnson's ELMA.

\begin{table}[h]
\centering
\caption{QF8 vs.\ Johnson's ELMA.}
\label{tab:vs-johnson}
\begin{tabular}{@{}p{3.5cm}p{5cm}p{5cm}@{}}
\toprule
Feature & Johnson ELMA (8-bit) & QF8 \\
\midrule
Log fractional bits  & $\sim$6 (posit tapered)     & 4 (fixed \texttt{u3.4})     \\
Encoding             & Posit regime bits (variable) & Fixed-width \texttt{u3.4}   \\
Block scaling        & None (self-scaled)           & E8M0 per 32 elements        \\
LUT size             & 64 entries ($2^6$)           & 16 entries ($2^4$) --- $4\times$ smaller \\
Optimality proof     & None                         & Minimax NMSE (Theorem~\ref{thm:minimax}) \\
\bottomrule
\end{tabular}
\end{table}

QF8 should be framed as building on Johnson's architecture with three improvements:
(a)~fixed-width encoding for hardware simplicity,
(b)~block scaling for precision/range separation,
(c)~formal minimax optimality analysis.

\subsection{Relationship to MX Formats (OCP Microscaling)}

QF8 borrows the E8M0 block scaling philosophy directly from the OCP MX specification
(Rouhani et al., 2023). The difference is the per-element encoding:
MXFP8 uses IEEE-like $4e + 3m$ (8 levels/octave); QF8 uses a log code (16 levels/octave).
\textbf{QF8 is ``MX block scaling + logarithmic per-element encoding.''}

\subsection{Relationship to NF4 (QLoRA)}

Both NF4 (Dettmers et al., 2023) and QF8 exploit distribution-aware encoding.
NF4 places levels at Gaussian quantiles (optimal for a specific distribution);
QF8's log-uniform spacing is minimax-optimal across all distributions.
NF4 is a storage format (dequantize before compute); QF8 is a computational format
(arithmetic in log domain).

\subsection{Logarithmic Number Systems}

QF8 is a descendant of the Logarithmic Number System (LNS), known since the 1970s
(Kingsbury \& Rayner, 1971; Swartzlander \& Alexopoulos, 1975).
The fundamental idea of log-domain multiplication is not new.
QF8's novelty is the specific combination of format design, block scaling, and formal
optimality analysis for ML workloads.

\subsection{What Is Genuinely Novel}

\begin{enumerate}[leftmargin=2em]
  \item \textbf{The equalization property (Theorem~\ref{thm:minimax}, Step 1).}
    The log-uniform quantizer achieves $\MSRE = \varepsilon^2/12$ for \emph{any} source density $f$.
    This density-independence follows from the exact $x^2$ cancellation in the NMSE ratio ---
    a structural property absent in all prior companding results.
    Smith~[15] observed amplitude-independent SNR for log-PCM, but his result is about signal level,
    not source distribution, and his distortion formula remains distribution-dependent.
  \item \textbf{Fixed-rate minimax optimality.}
    As a consequence of (1), log-uniform uniquely minimizes worst-case NMSE over all source densities.
    Sun \& Goyal~[16] prove log companding is optimal for variable-rate MSRE quantization but show
    it is \emph{not} generally optimal for fixed-rate. Theorem~\ref{thm:minimax} provides the
    complementary fixed-rate minimax result --- which holds precisely \emph{because} of (1).
    No prior 8-bit ML format has any proof of optimality.
  \item \textbf{The specific format:} \texttt{u3.4} log code + E8M0 block scaling + $O(1)$ bit-trick decode.
  \item \textbf{Quantified advantage:} $+6.6$~dB / $4.6\times$ NMSE improvement over FP8 E4M3 at
    identical storage cost ($8.25$ bits/element). The advantage is mathematical: it follows from
    $16$ vs.\ $8$ levels per octave.
  \item \textbf{Hardware cost analysis:} $3.5\times$ cheaper multiplier than FP8 ($66$ vs.\ $231$ gates).
\end{enumerate}

\subsection{The One-Sentence Pitch}

\begin{quote}
\emph{QF8 is what you get when you take Johnson's log-domain multiply architecture,
replace the posit tapered encoding with a simpler fixed-width log code,
add MX-style block scaling, and prove that the resulting format is minimax-optimal ---
yielding $2\times$ the precision per octave of FP8 at comparable hardware cost.}
\end{quote}


%% ═══════════════════════════════════════════════════════
\section{Future Work}
\label{sec:future}
%% ═══════════════════════════════════════════════════════

\subsection{Scaling Validation}

The TinyGPT-2 result (2~layers, 500 steps) demonstrates format viability but says little about
behavior at scale. Priority next steps:
\begin{enumerate}[leftmargin=2em]
  \item GPT-2 Small (124M parameters, 300k steps) --- does the advantage hold?
  \item GPT-2 Medium (355M) --- do any instabilities emerge?
  \item Comparison against MXFP8 with per-tensor scaling.
\end{enumerate}

\subsection{Gradient Quantization}

QF8's \texttt{u3.4} encoding gives 8~octaves per element, which may be insufficient for gradients
(which need wider dynamic range than weights/activations).
A \texttt{u4.3} variant (4~integer + 3~fractional bits = wider range, lower precision)
for backward passes is a natural extension but has not been validated.
FP8 training uses E5M2 for gradients for the same reason.


\subsection{BMD Decoder Conjecture}

\begin{conjecture}[BMD Decoder Characterization]
The class of monotone BMD decoders from $B$-bit integers to positive reals is exactly the class
of functions expressible as $\hat{D}(n) = \mathrm{float\_reinterpret}(a \cdot n + b)$
for integer constants $a, b$, or compositions of a bounded number of such reinterpretations
with integer arithmetic.
\end{conjecture}

This remains unproven and should be either proved rigorously or clearly labeled as a conjecture
in any paper.



%% ═══════════════════════════════════════════════════════
\section{Appendix: Rate-Distortion Context}
\label{sec:appendix-rd}
%% ═══════════════════════════════════════════════════════

\subsection{Gaussian Source Rate-Distortion}

For $X \sim \mathcal{N}(0, \sigma^2)$:
\[
  R(D) = \frac{1}{2}\log_2\!\left(\frac{\sigma^2}{D}\right), \quad D \leq \sigma^2
\]
This yields the \textbf{6 dB/bit rule}: each additional bit reduces MSE by $4\times$ (6.02 dB).

\subsection{Scalar Quantization Gap (Zador)}

The maximum improvement from scalar to infinite-dimensional vector quantization:
\[
  \frac{G_1}{G_\infty} = \frac{\pi e}{6} \approx 1.42 \quad (1.53 \text{ dB})
\]
This modest gap means a well-designed scalar quantizer like QF8 comes within 1.53~dB
of the VQ optimum.

\subsection{Bit Allocation}

For tensor types $t \in \{w, a, g\}$ with sensitivity weights $c_t$ and distribution
variances $\sigma_t^2$, the optimal allocation minimizes:
$\sum_t c_t \sigma_t^2 \cdot 2^{-2B_t}$ subject to $\sum_t B_t = B_{\text{total}}$.

\textbf{Solution (reverse water-filling):}
\[
  B_t^* = \frac{B_{\text{total}}}{|\mathcal{T}|} + \frac{1}{2}\log_2\!\left(\frac{c_t \sigma_t^2}{G}\right),
  \qquad G = \left(\prod_t c_t \sigma_t^2\right)^{1/|\mathcal{T}|}
\]

With corrected parameters ($c_w = 1, c_a = 1, c_g = 10^{-6}$), the near-zero gradient
allocation ($B_g = 0.18$) reflects that $c_g = \eta^2 \approx 10^{-6}$ dramatically
reduces gradient sensitivity, motivating stochastic rounding or shared-exponent compression
for gradients.


%% ═══════════════════════════════════════════════════════
\section*{References}
%% ═══════════════════════════════════════════════════════

\begin{enumerate}[label={[\arabic*]}, leftmargin=2.5em]
  \item Johnson, J. (2018). ``Rethinking Floating Point for Deep Learning.'' arXiv:1811.01721.
  \item Micikevicius, P. et al. (2022). ``FP8 Formats for Deep Learning.'' arXiv:2209.05433.
  \item Rouhani, B. et al. (2023). ``Microscaling Data Formats for Deep Learning.'' arXiv:2310.10537.
  \item Dettmers, T. et al. (2023). ``QLoRA: Efficient Finetuning of Quantized LLMs.'' arXiv:2305.14314.
  \item Gustafson, J. \& Yonemoto, I. (2017). ``Beating Floating Point at its Own Game: Posit Arithmetic.''
  \item Horowitz, M. (2014). ``Computing's Energy Problem (and what we can do about it).'' ISSCC.
  \item Jouppi, N. et al. (2017). ``In-Datacenter Performance Analysis of a Tensor Processing Unit.'' ISCA.
  \item Kingsbury, N. \& Rayner, P. (1971). ``Digital filtering using logarithmic arithmetic.'' Electronics Letters.
  \item Swartzlander, E. \& Alexopoulos, A. (1975). ``The Sign/Logarithm Number System.'' IEEE Trans.\ Computers.
  \item Miyashita, D. et al. (2016). ``Convolutional Neural Networks using Logarithmic Data Representation.'' arXiv:1603.01025.
  \item Sze, V. et al. (2017). ``Efficient Processing of Deep Neural Networks: A Tutorial and Survey.'' Proc.\ IEEE.
  \item Kuzmin, A., Van Baalen, M., Ren, Y., Nagel, M., Peters, J., Blankevoort, T. (2022). ``FP8 Quantization: The Power of the Exponent.'' arXiv:2208.09225.
  \item Bennett, W.R. (1948). ``Spectra of Quantized Signals.'' \textit{Bell System Technical Journal} 27(3):446--472.
  \item Panter, P.F. and Dite, W. (1951). ``Quantization Distortion in Pulse-Count Modulation with Nonuniform Spacing of Levels.'' \textit{Proceedings of the IRE} 39(1):44--48.
  \item Smith, B. (1957). ``Instantaneous Companding of Quantized Signals.'' \textit{Bell System Technical Journal} 36(3):653--709.
  \item Sun, J.Z. and Goyal, V.K. (2011). ``Scalar Quantization for Relative Error.'' \textit{Proc.\ IEEE Data Compression Conference (DCC)}, pp.~293--302.
  \item Gray, R.M. and Neuhoff, D.L. (1998). ``Quantization.'' \textit{IEEE Trans.\ Information Theory} 44(6):2325--2383.
  \item Gersho, A. and Gray, R.M. (1992). \textit{Vector Quantization and Signal Compression}. Kluwer Academic.
\end{enumerate}

\end{document}
